\bfseries Equivariant convolution-based functions on tensor fields

### 2022-07-27 paper/main.tex
content

\section{Problem setup}\label{sec:setup}

For our purposes, a cosmological simulation is a set of $N^3$ particles $n$ with $1\leq n\leq N^3$, each of which has a (scalar) mass $m_n$, a (vector) position $q_n$, and a (vector) veloctiy $v_n$, and each of which is a function of time (or computed or stored at a set of times $t$).
The positions $q_n$ live in a cubical 3-torus of side length $L$.
In the dark-matter-only case, the simulation dynamics depend only on the masses, positions, velocities $\{m_n, q_n, v_n\}_{n=1}^N$, time $t$, Newton's constant $G$, and (hopefully weakly) the number $N^3$ and length $L$.

Because the simulations are designed to describe the evolution of the Universe, the particles begin in a nearly-uniform density state, usually very close to a uniform, torus-filling, cubical lattice.
The particles begin at a very early time $t$ with tiny (vector) offsets $\Delta q_n$ away from lattice sites $q^{(0)}_n$ such that $q_n = q^{(0)}_n + \Delta q_n$, and tiny (vector) velocities $v_n$.
In detail these positional offsets and velocities are closely related by something like a Zeldovich approximation.

The uniform, torus-filling, cubical lattice of $N^3$ points $q^{(0)}_n$ can be thought of as a 3-dimensional ``image''.
Each particle can be thought of as a ``pixel'' in that image.
At each snapshot of the simulation at time $t$, each pixel has an associated (vector) positional offset $\Delta q_n$, (vector) velocity $v_n$, and a few scalars ($m_n, t, G$ at least).
In addition to these native scalars, we can produce at each pixel $n$ three additional scalars $q_n^\top q_n, q_n^\top v_n, v_n^\top v_n$, which are the inner products of the vectors with themselves and each other.
At late times, the simulation shows almost no memory of the initial uniform, torus-filling, cubical lattice.
However, each particle has an identity $n$ that can be associated with its original lattice location $q^{(0)}_n$, so its properties can be thought of as being encoded in that original lattice-location pixel.
That is, this is a problem of scalars and vectors on a uniform, cubical lattice in a cubical 3-torus.

There are some exact and approximate symmetries to consider in this problem.
The approximate symmetries are the fundamental symmetries of classical physics, which include translation, rotation, and parity.
These symmetries are only approximate in the n-body simulation context, because there is a cubical 3-torus with an orientation and a size, and there is a cubical lattice inside that 3-torus.
Generators of the exact symmetries are the three torus-axis-aligned spatial rotations by $\pi/2$ and one reflection across one of the three axes of the torus.

@@

We consider the orthogonal group $O(d)$ and $g\in O(d)$. For our purposes $d$ will be 2 or 3, but everything generalizes.  
We consider an image $A$ in a cubical $d$-torus with $N$ equally spaced pixels in each dimension ($N^d$ pixels total). Each pixel can contain tensors and pseudotensors (of which scalars, pseudoscalars, vectors, and pseudovectors are particular cases) that we define below.
Similarly, each linear geometric convolution operator $C$ will be a pixelized cubical image patch with $M$ pixels (with $M$ odd) in each dimension ($M^d$ pixels total) with a tensor or pseudotensor in each pixel, which acts as a linear pixel weight.

@@

\begin{definition}[vectors, $k$-tensors, and the world]
Let $O(d)$ act on vectors in $\mathbb R^d$ via the canonical action.
It also acts on lists of vectors $(\mathbb R^d)^n$ by acting on each of the $n$ vectors by the same action.
It also acts on rank-1 $k$-tensors in $(\mathbb R^d)^{\otimes k}$ by $g\cdot (v_{1}\otimes\ldots \otimes v_k) = (g\cdot v_1)\otimes \ldots \otimes (g\cdot v_k)$ and extending linearly to higher-rank tensors made from sums of rank-1 tensors.
Let $W$ (aka ``the world'') be the comprehensive set of all vectors and tensors that exist.
Note that $O(d)$ can be seen as acting on the world.
\end{definition}

@@

\begin{definition}[pseudovector and $k$-pseudotensor]
SOLE CAN WE SAY HERE that these are ``like a vector and like a tensor'' but there is an additonal $\det(g)$ coming in?
\end{definition}

\begin{definition}[scalar and pseudoscalar]
We say that $a\in \mathbb R$ is a $0$-tensor or scalar iff $a$ is an invariant of the action.
Namely, $a(W) = a(g\cdot W)$ for all $g\in O(d)$.
We say that
$a\in \mathbb R$ is a $0$-pseudotensor or pseudoscalar iff $a(g \cdot W) = \det(g)\,a(W)$ for all $g\in O(d)$,
where $\det(g)$ is the determinant of the matrix representation $M(g)$ of $g$.
\end{definition}

>>>>

@@

We consider the orthogonal group $O(d)$ and $g\in O(d)$. For our purposes $d$ will be 2 or 3, but everything generalizes.  
We consider an image $A$ in a $d$-torus with $N$ equally spaced pixels in each dimension ($N^d$ pixels total). Each pixel can contain tensors and pseudotensors (of which scalars, pseudoscalars, vectors, and pseudovectors are particular cases) that we define below.
Similarly, each linear geometric convolution operator $C$ will be a pixelized cubical image patch with $M$ pixels (with $M$ odd) in each dimension ($M^d$ pixels total) with a tensor or pseudotensor in each pixel, which acts as a linear pixel weight.

@@

\begin{definition}[$O(d)$ action on vectors and $k$-tensors] \label{def.tensors}
Let $O(d)$ act on vectors in $\mathbb R^d$ via the canonical action.
It also acts on rank-1 $k$-tensors in $(\mathbb R^d)^{\otimes k}$ by $g\cdot (v_{1}\otimes\ldots \otimes v_k) = (g\cdot v_1)\otimes \ldots \otimes (g\cdot v_k)$ and extending linearly to higher-rank tensors made from sums of rank-1 tensors.

It also acts on lists of $k$-tensors $((\mathbb R^d)^{\otimes k})^n$ by acting on each of the $n$ $k$-tensors by the same action.
%Let $W$ (aka ``the world'') be the comprehensive set of all vectors and tensors that exist.
%Note that $O(d)$ can be seen as acting on the world.
\end{definition}

@@

\begin{definition}[scalars and pseudoscalars]
We say that $a\in \mathbb R$ is a $0$-tensor or scalar iff $a$ is an invariant of the action.
Namely, $a(W) = a(g\cdot W)$ for all $g\in O(d)$.
We say that
$a\in \mathbb R$ is a $0$-pseudotensor or pseudoscalar iff $a(g \cdot W) = \det(g)\,a(W)$ for all $g\in O(d)$,
where $\det(g)$ is the determinant of the matrix representation $M(g)$ of $g$.
\end{definition}

\begin{definition}[pseudovectors and $k$-pseudotensors]
We say that $T\in (\mathbb R^d)^{\otimes k}$ if a $k$-pseudotensor if it is acted by $O(d)$ by the following action: $g\odot T = \det(g)\, g \cdot T$ where $g\cdot T$ is the cannonical action on tensors as in Definition \ref{def.tensors}.
\end{definition}
